Quantum Interval-Valued Probability: Contextuality and the Born Rule

### 2017-09-04 prop_letter.tex
prop_letter.tex \section{Gleason} Updated by Gerardo...
sec 4

\noindent The definition puts no restrictions on the choice of
intervals, only on the assignment of intervals to events. When
$\delta=0$, every event must be assigned a probability in either
interval $\imposs$ or $\necess$, i.e., every event is completely
determined with certainty. As $\delta$ gets larger, the QIVPM allows
for more indeterminate behavior.

The expectation value of an observable relative to a 0-deterministic
QIVPM is fully determinate and is equal to one of the eigenvalues of
that observable. To see this, note that given an orthonormal basis
$\Omega=\{ \ket{\psi_0},\ket{\psi_1},\ldots\}$, a 0-deterministic
QIVPM must map exactly one of the projectors $\proj{\psi_i}$ to
$\necess$ and all others to $\imposs$. This is because by
Eq.~(\ref{eq:QIVPM-constraints}b), we have
$\bar{\mu}(\Sigma_i{\proj{\psi_i}})=\necess$ and by
Eq.~\ref{eq:QuantumInterval-valuedProbability-Inclusion}, we must have
one of the $\bar{\mu}(\proj{\psi_i})=\necess$ and all others mapped to
$\imposs$. Given any state~$\rho$ that is consistent with this QIVPM
on all the projectors in $\Omega$, we have by Eq.~(\ref{eq:hbar}) that
$\mu_{\rho}^{\mathrm{B}}$ must also map exactly one of the projectors
to 1 and all others to 0. If an observable has a spectral
decomposition along $\Omega$ then, by
Eq.~(\ref{eq:quantum-expectation}), its expectation value relative to
$\mu_{\rho}^{\mathrm{B}}$ is the eigenvalue $\lambda$ whose projector
is mapped to 1. It therefore follows, by
Eq.~(\ref{eq:quantum-interval-expectation-core}), that the expectation
value relative to the 0-deterministic~${\bar{\mu}}$ is fully
determinate and lies in the interval $[\lambda,\lambda]$.

@@

\noindent To explain why this result is equivalent to the original
Kochen-Specker theorem and to prove it at the same time, we proceed by
assuming a 0-deterministic QIVPM $\bar{\mu}$ and derive a
contradiction to the original Kochen-Specker theorem. Our
counterexample uses the simpler proof for a Hilbert space of dimension
$d=4$ and is constructed as follows.

@@

\noindent The observables are constructed using the Pauli matrices
$\left\{ \mathbb{1},\sigma_{x},\sigma_{y},\sigma_{z}\right\}$ whose
eigenvalues are all either 1 or -1. They are arranged such that in
each row and column, \emph{except the column $j=2$}, every observable
is the product of the other two. In the $j=2$ column, we have instead
that
$\left(\sigma_{z}\otimes\sigma_{z}\right)\left(\sigma_{x}\otimes\sigma_{x}\right)=-\sigma_{y}\otimes\sigma_{y}$. Now
assume a QIVPM ${\bar{\mu}}$; the expectation values of the
observables in each row relative to this QIVPM are fully determinate
and must lie in either the interval $[1,1]$ or the interval $[-1,-1]$
depending on which eigenvalue is the one whose associated projector is
certain. Since the product of any two observables in a row is equal to
the third, there must be an even number of occurrences of the interval
$[-1,-1]$ in each row and hence in the entire table. However, looking
at the expectation values of the observables in each column, there
must be an even number of occurrences of the interval $[-1,-1]$ in the
first two columns and an odd number in the $j=2$ column and hence in
the entire table. The contradiction implies the non-existence of the
assumed QIVPM.

@@

% A 0-deterministic QIVPM~$\bar{\mu}$ has an important property
% regarding to the expectation values of a family of $M$ commuting
% observables~$\left\{ \mathbf{O}_{j}\right\} _{j=1}^{M}$ relative to
% $\bar{\mu}$. Since $\left\{ \mathbf{O}_{j}\right\} _{j=1}^{M}$
% commutes, they can be diagonalized by a common orthonormal
% basis~$\Omega=\left\{ \ket{k}\right\} _{k=0}^{d-1}$.  Since
% $\bar{\mu}$ is 0-deterministic, a unique $\ket{k_{0}}$ and a unique
% quantum probability measure~$\mu$ exists such that
% $\bar{\mu}\left(P\right)=\left\{ \mu\left(P\right)\right\} $ and
% $\mu\left(P\right)\in\left\{ 0,1\right\} $ for all $P\in\events$;
% $\mu\left(\proj{k}\right)=1$ if and only if
% $k=k_{0}$~\citep{HOSTunpublished}.  If we denote the product
% observables~$\prod_{j=1}^{M}\mathbf{O}_{j}$ as
% $\overline{\mathbf{O}}$, the expectation values of the
% observables~$\expval{\mathbf{O}_{j}}_{\bar{\mu}}$ and
% $\expval{\overline{\mathbf{O}}}_{\bar{\mu}}$ are equal to the
% singleton sets whose only elements are $\expval{\mathbf{O}_{j}}_{\mu}$
% and $\expval{\overline{\mathbf{O}}}_{\mu}$, respectively. Both
% $\prod_{j=1}^{M}\expval{\mathbf{O}_{j}}_{\mu}$ and
% $\expval{\overline{\mathbf{O}}}_{\mu}$ are equal to the product of the
% eigenvalues corresponding to the pure state~$\ket{k_{0}}$ for each
% observable~$\mathbf{O}_{j}$. By adopting an intuitive product
% rule~$\left\{ x_{0}\right\} \cdot\left\{ x_{1}\right\} =\left\{
%   x_{0}x_{1}\right\} $, we have the following property:
% \begin{equation}
% \prod_{j=1}^{M}\expval{\mathbf{O}_{j}}_{\bar{\mu}}=\expval{\prod_{j=1}^{M}\mathbf{O}_{j}}_{\bar{\mu}}\,.\label{eq:Kochen-Specker-Rules}
% \end{equation}

@@

When $\delta \ge \frac{1}{3}$, the situation abruptly
changes. Assuming a Hilbert space $\Hilb$ of dimension $3$, then for
any projector $P$ that projects onto an $n$-dimensional subspace of
$\Hilb$, it is straightforward to check that
$\bar{\mu}\left(P\right)=\left[n/3,n/3\right]$ is a valid
$\frac{1}{3}$-deterministic QIVPM. In other words, when the uncertainty
in measurements becomes so large, the Kochen-Specker theorem no longer
holds and it becomes possible to map every observable to some (quite
inaccurate) probability interval.

>>>>

\noindent The definition puts no restrictions on the choice of
intervals, only on the assignment of intervals to events. When
$\delta=0$, every event must be assigned a probability in either
interval $\imposs$ or $\necess$, i.e., every event is completely
determined with certainty. As $\delta$ gets larger, the QIVPM allows
for more indeterminate behavior. For example, assuming a set of
intervals including $\imposs$, $\necess$, and
$\{ [0,1/4], [1/4,1/2], [1/2,3/4], [3/4,1] \}$, the probability
measure that maps every event to either $\imposs$ or $\necess$ would
be a 0-deterministic QIVPM, and the probability measure that maps
every event to one of $\imposs$, $\necess$, $[0,1/5]$, or $[4/5,1]$
would be a $\delta$-deterministic QIVPM for every
$1/5 \le \delta \le 1$.

The expectation value of an observable relative to a 0-deterministic
QIVPM is fully determinate and is equal to one of the eigenvalues of
that observable. To see this, note that given an orthonormal basis
$\Omega=\{ \ket{\psi_0},\ket{\psi_1},\ldots\}$, a 0-deterministic
QIVPM must map exactly one of the projectors $\proj{\psi_i}$ to
$\necess$ and all others to $\imposs$. This is because by
Eq.~(\ref{eq:QIVPM-constraints}b), we have
$\bar{\mu}(\Sigma_i{\proj{\psi_i}})=\necess$ and by
Eq.~\ref{eq:QuantumInterval-valuedProbability-Inclusion}, we must have
one of the $\bar{\mu}(\proj{\psi_i})=\necess$ and all others mapped to
$\imposs$. Given any state~$\rho$ that is consistent with this QIVPM
on all the projectors in $\Omega$, we have by Eq.~(\ref{eq:hbar}) that
$\mu_{\rho}^{\mathrm{B}}$ must also map exactly one of the projectors
in $\Omega$ to 1 and all others to 0. If an observable has a spectral
decomposition along $\Omega$ then, by
Eq.~(\ref{eq:quantum-expectation}), its expectation value relative to
$\mu_{\rho}^{\mathrm{B}}$ is the eigenvalue $\lambda$ whose projector
is mapped to 1. It therefore follows, by
Eq.~(\ref{eq:quantum-interval-expectation-core}), that the expectation
value relative to the 0-deterministic~${\bar{\mu}}$ is fully
determinate and lies in the interval $[\lambda,\lambda]$.

@@

\noindent To explain why this result is equivalent to the original
Kochen-Specker theorem and to prove it at the same time, we proceed by
assuming a 0-deterministic QIVPM $\bar{\mu}$ and derive a
contradiction to the original Kochen-Specker theorem. Instead of
adapting the more complicated proof for $d=3$, our counterexample uses
the simpler proof for a Hilbert space of dimension $d=4$ and is
constructed as follows.

@@

\noindent The observables are constructed using the Pauli matrices
$\left\{ \mathbb{1},\sigma_{x},\sigma_{y},\sigma_{z}\right\}$ whose
eigenvalues are all either 1 or -1. They are arranged such that in
each row and column, \emph{except the column $j=2$}, every observable
is the product of the other two. In the $j=2$ column, we have instead
that
$\left(\sigma_{z}\otimes\sigma_{z}\right)\left(\sigma_{x}\otimes\sigma_{x}\right)=-\sigma_{y}\otimes\sigma_{y}$. Now
assume a 0-deterministic QIVPM ${\bar{\mu}}$; the expectation values
of the observables in each row relative to this 0-deterministic QIVPM
are fully determinate and must lie in either the interval $[1,1]$ or
the interval $[-1,-1]$ depending on which eigenvalue is the one whose
associated projector is certain. Since the product of any two
observables in a row is equal to the third, there must be an even
number of occurrences of the interval $[-1,-1]$ in each row and hence
in the entire table. However, looking at the expectation values of the
observables in each column, there must be an even number of
occurrences of the interval $[-1,-1]$ in the first two columns and an
odd number in the $j=2$ column and hence in the entire table. The
contradiction implies the non-existence of the assumed 0-deterministic
QIVPM. We therefore prove a stronger statement of contextuality that
includes the effects of finite-precision.

@@

% A 0-deterministic QIVPM~$\bar{\mu}$ has an important property
% regarding to the expectation values of a family of $M$ commuting
% observables~$\left\{ \mathbf{O}_{j}\right\} _{j=1}^{M}$ relative to
% $\bar{\mu}$. Since $\left\{ \mathbf{O}_{j}\right\} _{j=1}^{M}$
% commutes, they can be diagonalized by a common orthonormal
% basis~$\Omega=\left\{ \ket{k}\right\} _{k=0}^{d-1}$.  Since
% $\bar{\mu}$ is 0-deterministic, a unique $\ket{k_{0}}$ and a unique
% quantum probability measure~$\mu$ exists such that
% $\bar{\mu}\left(P\right)=\left\{ \mu\left(P\right)\right\} $ and
% $\mu\left(P\right)\in\left\{ 0,1\right\} $ for all $P\in\events$;
% $\mu\left(\proj{k}\right)=1$ if and only if
% $k=k_{0}$~\citep{HOSTunpublished}.  If we denote the product
% observables~$\prod_{j=1}^{M}\mathbf{O}_{j}$ as
% $\overline{\mathbf{O}}$, the expectation values of the
% observables~$\expval{\mathbf{O}_{j}}_{\bar{\mu}}$ and
% $\expval{\overline{\mathbf{O}}}_{\bar{\mu}}$ are equal to the
% singleton sets whose only elements are $\expval{\mathbf{O}_{j}}_{\mu}$
% and $\expval{\overline{\mathbf{O}}}_{\mu}$, respectively. Both
% $\prod_{j=1}^{M}\expval{\mathbf{O}_{j}}_{\mu}$ and
% $\expval{\overline{\mathbf{O}}}_{\mu}$ are equal to the product of the
% eigenvalues corresponding to the pure state~$\ket{k_{0}}$ for each
% observable~$\mathbf{O}_{j}$. By adopting an intuitive product
% rule~$\left\{ x_{0}\right\} \cdot\left\{ x_{1}\right\} =\left\{
%   x_{0}x_{1}\right\} $, we have the following property:
% \begin{equation}
% \prod_{j=1}^{M}\expval{\mathbf{O}_{j}}_{\bar{\mu}}=\expval{\prod_{j=1}^{M}\mathbF{O}_{j}}_{\bar{\mu}}\,.\label{eq:Kochen-Specker-Rules}
% \end{equation}

@@

When $\delta \ge \frac{1}{3}$, the situation abruptly
changes. Assuming a Hilbert space $\Hilb$ of dimension $3$, then for
any projector $P$ that projects onto an $n$-dimensional subspace of
$\Hilb$, it is straightforward to check that
$\bar{\mu}\left(P\right)=\left[n/3,n/3\right]$ is a valid
$\frac{1}{3}$-deterministic QIVPM. In other words, when the
uncertainty in measurements becomes so large, the Kochen-Specker
theorem no longer holds and it becomes possible to map every
observable to some (quite inaccurate) probability interval, and hence
the theory becomes non-contextual.
